Basics of Hardware Cache Optimization

### 2017-05-28 tex/basic-concepts.tex
Split basic-concepts.tex subsections into files

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

% The reason why the loops take the same amount of time has to do with memory.  The
% running time of these loops is dominated by the memory accesses to the array, not by the
% integer multiplications.
%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
This is because the run times are \alts{primarily due to, dominated by} memory accesses.
Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
are 128 bytes apart,%
\footnote{16 \texttt{int64\_t} values of 8 bytes each}
so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
and so on~\cite[cf.][example 2]{gallery}.

\begin{listing}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{listing}

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

\subsection{Prefetching}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching
\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

\subsection{Associativity}
% https://en.wikipedia.org/wiki/CPU_cache#Associativity

\subsection{Cache Hits and Misses}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?

\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[2]{afmh}.

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
% summing the array.

\subsection{Temporal Locality}
TODO~\cite{drepper2007}.

\subsection{Memory Access Pattern}
\label{sec:map}

>>>>

\input{tex/basic-concepts/lines.tex}
\input{tex/basic-concepts/prefetching.tex}
\input{tex/basic-concepts/assoc.tex}
\subsection{Cache Hits and Misses}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
\input{tex/basic-concepts/spat_loc.tex}
\input{tex/basic-concepts/temp_loc.tex}
\subsection{Memory Access Pattern}
\label{sec:map}
